Introducing an Ecological Trait-data Standard

### 2017-10-27 manuscript/manuscript.Rmd
start revision of manuscript, add references, include comments of Klaus and Catrin

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas
  Ostrowski, Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @weiner14] and parameterize ecological population-dynamic models [@refs].

On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @simons16;@hopfenmuller14; @mouillot13;  @vandewalle10; @borschig13]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].

In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.  
Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.

@@

Despite the obvious benefit, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits heterogeneous. 
Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme. Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.

Most prominently, plant traits from many different sources -- including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08] -- have been collated into the quickly growing TRY database [@kattge11].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism  -->

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/), or the Carabids.org web portal (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @gossner15, @gossner15c, @renner14]. However, a unified platform and harmonizing scheme for the datasets is still lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits. [@refs]

@@

## Need for ontologies for data standardisation

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15]. However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits are compliant with this demand for open data. <!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].--> Open-access publications in low-threshold data repositories such as Figshare, datadryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

<!--Today, several general biodiversity data intitiatives are striving for a global integration of ecological and biological information. Only few so far developed universal data frameworks for trait data, such as the EOL TraitBank, which is the most general framework to date (Parr et al 2015). TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. -->
To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored.  Such a general data structure for quantitative trait data has been developed within the Encyclopedia of Life Project for upload of data into TraitBank [@parr16], relying on definitions of terms provided by the Darwin Core Standard (DWC, ...). 
However, the Terms of the Darwin Core Standard and available Extensions do not cover all of the aspects required in the various use-cases of trait-based research.

@@

With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

We propose a versatile data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. In these 'occurrence data tables', each row is centered around a single physical occurrence of a species. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.

Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one measured value.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14a]. They further draw from a defined set of columns, which makes it easier to merge datasets.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets. Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured. a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e. one physical occurence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.   }

@@

## common terms for trait data

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset compliant with the EST, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above (Table 1).
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).  The full list of terms  can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).

A dataset that applies these terms should relate to the EST in its metadata. Since the traitdata standard is a resource under open community development, the metadata also should refer to a version number.  This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g. http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).

@@

| Term | valueType | vocabulary | Identifier | Definition | 
|------|-----------|------------|------------|------------|
|scientificName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificname |Original character string provided as species name by the data owner (kept for reference only)| 
|taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected. Each entry should link to a published reference species list/taxonomy. Avoid synonyms.|
|taxonRank | factor | domain; kingdom; subkingdom; superhpylum [...] |http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonrank | The taxonomic rank of the most specific name in the scientificName. Recommended best practice is to use a controlled vocabulary.|
|kingdom | factor | |http://fdschneider.de/bexis_traits/traitdatastandard.html#kingdom | The full scientific name of the kingdom in which the taxon is classified.| 
|traitName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#traitname |Name of the trait as used by the data provider (kept for compatibility on provider side).|
|traitID|factor|see traitlist|http://fdschneider.de/bexis_traits/traitdatastandard.html#traitid ||
|traitValue|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvalue |Measured raw value or factor level for this species trait as measured and provided by the author.|
|traitUnit|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunit |Reports the unit that the author's raw data were measured in, if applicable (only for numeric values).|
|scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive name of the trait reported. This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
|traitValueStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvaluestd |Standardised measured value or factor level for this species trait. Numerical values must use the correct unit. Factor levels must be compliant with the thesaurus of traits.|
|traitUnitStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunitstd |The units associated with the measurementValue. The expected standard unit for this trait as defined in the traitlist/ thesaurus.|
|measurementID|	character||	http://fdschneider.de/bexis_traits/traitdatastandard.html#measurementid |A unique, dataset-specific identifier. Links multi-value trait measurements, e.g. x-y-z coordinates of a morphometric landmark, biochemical compound quantities for different chainlengths. In this case, the trait names must specifiy the sub-measurement, e.g. "landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurenceid|A character string user defined by the data author. It is recommended best practice to use a globally unique identifier, if available, e.g. a museum ID referring to the individual specimen from which the data were measured. Usually does not apply for literature data.|
|warnings|character|	|	|Warnings from the R-script will be stored here, e.g. regarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table. User defined warnings and flags can be added as well, e.g. 'NOTUSE'.|

@@

## core elements of a trait-dataset

@@

Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11]. This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') . 
Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

@@

A similar approach to standardised trait names (`traitNameStd`) and harmonised reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
Resources for semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13]. However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.

### Defining trait thesauri and ontologies

@@

### relating core data to other datasets

>>>>

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  word_document:
    keep_md: yes
  pdf_document: default
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions (@refs).

<!-- more refs:  Reich et al., 1997; Wright et al., 2004; Chave et al., 2009; Schrodt et al., 2015-->

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].   
On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel17; @lajoje15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].  
In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->

@@

Despite these obvious benefit and grand prospects, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous. 
Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme. Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.

Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g. including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism -->

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], or the Carabids.org web portal (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications over the decades [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> ]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

@@

## Need for ontologies to enable data standardisation

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15]. However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits comply with this demand for open data. While it seems clear that factual data, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@ref], there is still no broad consensus achieved in the scientific community whether such data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship. <!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. legal standpoint in most countries: descriptive data are not copyright protected. Publicly fundet research must provide products. is still debated by the scientific community. -->

Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Although open for manual download, data on those platforms might be stored in proprietary (e.g. .xlsx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious or impossible. Since the platforms are open for any type of scientific data, no data standards are enforced upon the uploaded material. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.

Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12) and relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies  [@parr15]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities. The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered. To date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots -->

@@

With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the work of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

We define a versatile, minimal data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

With this framework, we build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12]. We further expand the definitions of these terms for the use case of trait data and add terms which cover the special demands of trait-based research, while maintaining semantic compatibility with existing initatives.

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix (a.k.a. wide-table format), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. In these occurrence data tables, each row is centered around a single physical occurrence of a species. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.

Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one measured value. The width of the table varies with the number of traits included. Merging data from different sources requires manual harmonization of these structures.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) which is assigned to a specific taxon. This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14a]. They further draw from a defined set of columns, which makes it easier to merge datasets.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets. Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured. a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e. one physical occurrence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.   }

@@

## Common terms for trait data

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above.
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata). 
A second set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail). 
The full glossary of terms can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication). Were possible, we relate these terms to existing terms in ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core ().

A dataset that applies these terms should refer to the data standard in its metadata. Since the traitdata standard is a resource under open community development, this reference also should include the version number. This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g. http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).

@@

| Term | valueType | vocabulary | Identifier | Definition | 
|------|-----------|------------|------------|------------|
|scientificName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificname |Original character string provided as species name by the data owner (kept for reference only)| 
|taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected. Each entry should link to a published reference species list/taxonomy. Avoid synonyms.|
|taxonRank | factor | domain; kingdom; subkingdom; superhpylum [...] |http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonrank | The taxonomic rank of the most specific name in the scientificName. Recommended best practice is to use a controlled vocabulary.|
|kingdom | factor | |http://fdschneider.de/bexis_traits/traitdatastandard.html#kingdom | The full scientific name of the kingdom in which the taxon is classified.| 
|traitName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#traitname |Name of the trait as used by the data provider (kept for compatibility on provider side).|
|traitID|factor|see traitlist|http://fdschneider.de/bexis_traits/traitdatastandard.html#traitid ||
|traitValue|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvalue |Measured raw value or factor level for this species trait as measured and provided by the author.|
|traitUnit|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunit |Reports the unit that the author's raw data were measured in, if applicable (only for numeric values).|
|scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive name of the trait reported. This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
|traitValueStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvaluestd |Standardised measured value or factor level for this species trait. Numerical values must use the correct unit. Factor levels must be compliant with the thesaurus of traits.|
|traitUnitStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunitstd |The units associated with the measurementValue. The expected standard unit for this trait as defined in the traitlist/ thesaurus.|
|measurementID|	character||	http://fdschneider.de/bexis_traits/traitdatastandard.html#measurementid |A unique, dataset-specific identifier. Links multi-value trait measurements, e.g. x-y-z coordinates of a morphometric landmark, biochemical compound quantities for different chainlengths. In this case, the trait names must specifiy the sub-measurement, e.g. "landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurrenceid|A character string user defined by the data author. It is recommended best practice to use a globally unique identifier, if available, e.g. a museum ID referring to the individual specimen from which the data were measured. Usually does not apply for literature data.|
|warnings|character|	|	|Warnings from the R-script will be stored here, e.g. regarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table. User defined warnings and flags can be added as well, e.g. 'NOTUSE'.|

@@

### core elements of a trait-dataset

@@

Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11]. This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'). 
Keeping a reference to a kingdom and class (using field `kingdom`, `class` as described by Darwin Core Standard and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15]. 
Alternatively to expanding the number of columns, these data could be contained in a separate data file which relates to the core dataset via the field `ScientificName`.

@@

A similar approach to standardised trait names and harmonised reported values in compliance with published ontologies is advised. Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI) of trait definitions or a measurement methodology.

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
To date, semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples). However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.

<!--### Defining own trait thesauri and ontologies-->

@@
